The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2013-10-29 paper/ms.tex
Comments about new plots
corrected some typos

The first quantitative conclusion of the effect of rotation in the
\ly line is that double peaks broaden and reduce their intensity
while the line center rises. This produces the impression that, as the
rotational velocity increases, the double peaks are merged into a
single broad emission peak. This is most evident at the highest
rotational velocities for the homogeneously distributed sources
(right panel in Figure \ref{fig:differentvelocities}).

@@

The transition to a single peaked line seems to occur in systems
where it is easier for the bulk of the photons to escape with the lowest
number of scatterings possible, allowing them not to move very far
from the center of the line. This might explain how the single peak stage
is easily achieved in the homogeneous source distribution. In this
case there is a fraction of the photons that are inside a photosphere
region with $\tau_{H,r}\ll \tau_{H}$ where $\tau_{H,r}$ is the optical
depth from the radius of emission to the sphere's surface. This
conditions allows the photons to escape with much less scatterings
compared to the photons emitted at the very center of the sphere. In
turn, it gives the photons less scatterings to be placed far from the
line center.

@@

The number of \ly photon scatterings affects its final frequency after
escaping the galaxy. In static gas geometries, a large value of the
optical depth correlates with a high number of scatterings, increasing
the probability of finding a \ly photon far from the center of the
line. In these cases the peak maxima shift away form the line center as
the amount of neutral hydrogen increases.

@@

@@

Figure \ref{fig:Nscatt2D} supports our hypothesis about
the photosphere in the homogeneous distribution. In that case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This explains the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, on the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

\begin{figure}
  \includegraphics[width=0.45\textwidth]{Neufeld.png}
 \caption{Escape fraction as a function of the
   product $(a\tau_{\rm H})^{1/3}\tau_{a}$. The analytic solution for
   the infinite is slab shown as a continuous line. Different dashed
   lines correspond to different rotational velocities.
   \label{fig:efvsNeufeld}}   
\end{figure}

The average number of scatterings is also reflected in the amount of
photons absorbed by dust. Under this considerations we do not expect
any change in the escape fraction for a dusty cloud with central
source of radiation given that the number of scatterings remains close
to constant. On the other hand, in the case of an homogeneous
radiation source the number of scatterings drops as $V_{\rm max}$
increases, which has to be reflected as an increasing escape fraction.

Figure \ref{fig:efvsv} shows the dependence of the escape fraction as
a function of the maximum rotational velocity, confirming our expectations. For the central source
distribution the escape fraction barely shows any change, while for
the homogeneous case there is a clear rise in the escape fraction for
high rotational velocities.  Rotation has a higher relative impact in
the models with low optical depth $\tau_{\rm H}=10^5$,$10^{6}$, where
it can rise from $(0.26, 0.28)$ ($V_{\rm}=0$\kms) up to $(0.37, 0.39)$ ($V_{\rm
  rot}=300$ \kms).

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends solely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. First of all we note that the escape fraction does not
increase from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, this is
explained because we are in a regime where the condition for the
analytic expectations  ($a\tau_{h}\gg 1$) does not hold. This figure
also shows that the escape fraction is higher by factors of $2-10$
than the expectation for a slab with centrally distributed
sources. Furthermore, the decrease of the escape fraction with
increasing optical depth is also slower.

\subsection{Off-Centered emission}

\begin{figure}
  \includegraphics[scale=0.3]{OF_spheres.png}
  \caption{The two solid circles show the two different spherical off-centered emitting
    regions. The angular velocity vector is defined to be along the
    $\hat{\i}$ direction and the observer is along the $\hat{\j}$ direction (where
    the reader is). 
    \label{fig:OCspheres}} 
\end{figure}

@@

For the Off-center emission Fig.\ref{fig:OCflux} the variations of 
$\mu$ increase as the rotational velocity increase, for $V_{\rm max}=(50kms^{-1}$
 $\bigtriangleup\mu \leq 7\%$ , for $V_{\rm max}=(100kms^{-1}$
$\bigtriangleup\mu\leq 12\%$ and finally for $V_{\rm max}=(200kms^{-1}$
$\bigtriangleup\mu\leq 15\%$.
%%%%%%%

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of this
 findings for the interpretation of available observational data. Our
 dicussion will focus on the qualitative aspects of diverse
 morpohological features observerd by \cite{Kulas12} ans \cite{Yamada2012}

We follow a classification of $Ly\alpha$ profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ that showed multiple peaks in their emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is a
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point. Group V presents
three peaks.

@@

Another relevant sample of lines was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation.

@@

In this paper, we have estimated the effects of gas bulk rotation on
the morphology of the Lyman $\alpha$ line in star forming
galaxies. The results are based on the study of an homogeneous sphere
rotating as a solid body. We explored  a range of models by varying
the rotation speed, hydrogen optical depth, dust optical depth and
initial distribution of Ly$\alpha$ photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

@@

We have quantified the results in different aspects. In Section
\ref{sec:widthpeak} we show that the line width increases up to a
factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition form a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed.

@@

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles and multi-peaked profiles could be explained by gas bulk 
rotation present in this LAEs.

>>>>

The first quantitative conclusion of the effect of rotation in the
\ly line is that double peaks broaden and reduce their intensity
while the line center rises. This gives the impression that, as the
rotational velocity increases, the double peaks are merged into a
single broad emission peak. This is most evident at the highest
rotational velocities for the homogeneously distributed sources
(right panel in Figure \ref{fig:differentvelocities}).

@@

The transition to a single peaked line seems to occur in systems
where it is easier for the bulk of the photons to escape with the lowest
number of scatterings possible, allowing them not to move very far
from the center of the line. This might explain how the single peak stage
is easily achieved in the homogeneous source distribution. In this
case there is a fraction of the photons that are inside a photosphere
region with $\tau_{H,r}\ll \tau_{H}$ where $\tau_{H,r}$ is the optical
depth from the radius of emission to the sphere's surface. These
conditions allows the photons to escape with much less scatterings
compared to the photons emitted at the very center of the sphere. In
turn, it gives the photons less scatterings to be placed far from the
line center.

@@

The number of \ly photon scatterings affects its final frequency after
escaping the galaxy. In static gas geometries, a large value of the
optical depth correlates with a high number of scatterings, increasing
the probability of finding a \ly photon far from the center of the
line. In these cases the peak maxima shifts away from the line center as
the amount of neutral hydrogen increases.

@@

%NEXT PARAGRAPH IS STRANGE

@@

Figure \ref{fig:Nscatt2D} supports our hypothesis about
the photosphere in the homogeneous distribution. In that case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This explains the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, in the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

\begin{figure}
  \includegraphics[width=0.45\textwidth]{Neufeld.png}
 \caption{Escape fraction as a function of the
   product $(a\tau_{\rm H})^{1/3}\tau_{a}$. The analytic solution for
   the infinite slab is shown as a continuous line. Different dashed
   lines correspond to different rotational velocities.
   \label{fig:efvsNeufeld}}   
\end{figure}

The average number of scatterings is also reflected in the amount of
photons absorbed by dust. Under these considerations we do not expect
any change in the escape fraction for a dusty cloud with central
source of radiation given that the number of scatterings remains close
to constant. On the other hand, in the case of a homogeneous
radiation source the number of scatterings drops as $V_{\rm max}$
increases, which has to be reflected as an increasing escape fraction.

Figure \ref{fig:efvsv} shows the dependence of the escape fraction as
a function of the maximum rotational velocity, confirming our expectations. For the central source
distribution the escape fraction barely shows any change, while for
the homogeneous case there is a clear increase in the escape fraction for
high rotational velocities.  Rotation has a higher relative impact in
the models with low optical depth $\tau_{\rm H}=10^5$,$10^{6}$, where
it can rise from $(0.26, 0.28)$ ($V_{\rm}=0$\kms) up to $(0.37, 0.39)$ ($V_{\rm
  rot}=300$ \kms).

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits where $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. First of all we note that the escape fraction does not
increase from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, this is
explained since we are in a regime where the condition for the
analytic expectations  ($a\tau_{h}\gg 1$) does not hold. This figure
also shows that the escape fraction is higher by factors of $2-10$
than the expectation for a slab with centrally distributed
sources. Furthermore, the decrease of the escape fraction with
increasing optical depth is also slower.

\subsection{Off-Centred emission}

\begin{figure}
  \includegraphics[scale=0.3]{OF_spheres.png}
  \caption{The two solid circles show the two different spherical off-centred emitting
    regions. The angular velocity vector is defined to be along the
    $\hat{\i}$ direction and the observer is along the $\hat{\j}$ direction (where
    the reader is). 
    \label{fig:OCspheres}} 
\end{figure}

@@

For the Off-center emission Fig.\ref{fig:OCflux} the variations of 
$\mu$ increase as the rotational velocity increase, for $V_{\rm max}=(50kms^{-1}$
 $\bigtriangleup\mu \leq 7\%$ , for $V_{\rm max}=(100kms^{-1}$
$\bigtriangleup\mu\leq 12\%$ and finally for $V_{\rm max}=(200kms^{-1}$
$\bigtriangleup\mu\leq 15\%$. This would imply that the amount of flux
we measure would depend in the angle between our line of sight and 
the rotation axis of the galaxy. 
%%%%%%%

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion will focus on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} ans \cite{Yamada2012}

We follow a classification of $Ly\alpha$ profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ that showed multiple peaks in their emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is an
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point. Group V presents
three peaks.

@@

Another relevant sample of lines was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-centre emission models with rotation.

@@

In this paper, we have estimated the effects of gas bulk rotation on
the morphology of the Lyman $\alpha$ line in star forming
galaxies. The results are based on the study of a homogeneous sphere
rotating as a solid body. We explored a range of models by varying
the rotation speed, hydrogen optical depth, dust optical depth and
initial distribution of Ly$\alpha$ photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

@@

We have quantified the results in different aspects. In Section
\ref{sec:widthpeak} we show that the line width increases up to a
factor of $2-3$ in comparison to the static case. Simultaneously, the
flux in the line center also increases with rotational velocity. This
increase induces the transition from a double peak to a single peak
line in the models where the radiation sources are homogeneously
distributed.

@@

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles and multi-peaked profiles could be explained by gas bulk 
rotation present in these LAEs.
